Leveraging The Progress in Automated Theorem Proving for Program Synthesis\\ \small{WWTF ICT 2022}

### 2022-03-28 WWTF/main.tex
asdf

\section{Introduction, background and state of the art (max. 2500 characters)}
%What is the topic and broad background of this research project? What is the state of the art in this field?
Programs are usually first written and then tested/verified.
In contrast, program synthesis allows a system developer to derive a program in one step.
At the same time, program synthesis holds the promise of freeing the programmer from low-level details and of enabling the programmer to focus on the high-level intent.
Program synthesis is a long-standing problem that crosses many areas of computer science and mathematics including programming languages, AI, formal methods and proof theory.
Traditionally, program synthesis has been formalized as a problem in deductive theorem proving~\cite{conf/ijcai/MannaW79}:
Given a logical formula that states that for all inputs there exists an output such that the required specification holds, one searches for a constructive proof of this formula and then extracts a program from this proof.
A classic route to program extraction is the \emph{Curry-Howard correspondence}~\cite{?} {\color{red}Standardreferenz Proofs and Types~\cite{girard1989proofs} oder Original Curry~\cite{curry1934functionality}} (also known as \emph{proofs-as-programs}) where one exploits the equivalence between proofs in intuitionist logic and functional programs in order to directly extract a functional program from an intuitionistic proof.
The Curry-Howard correspondence denotes a fundamental relationship between two seemingly unrelated formalisms, namely proofs systems on one hand and computational models on the other hand.
The Curry-Howard correspondence has been a major driving force for research spanning functional programming languages, type theory, proof systems and constructive logic.
However, progress on the deductive synthesis problem has been slow for many years because of a lack of automated and scalable provers.
Instead, a recent variant of the program synthesis problem has enabled dramatic progress (we refer to the special issue~\cite{fisman2022special} and the workshop\footnote{\url{https://simons.berkeley.edu/workshops/tfcs2021-1}} for an overview):
%\footnote{https://www.cs.technion.ac.il/~shaull/SYNT2022/index.html}
The logical specification is accompanied by a \emph{syntactic template} that constrains the space of possible implementations.
The synthesis procedure then searches for/enumerates candidate implementations and either rejects these candidates by finding violating input/output pairs or accept the generated program by verifying their correctness.
The syntactic approach is very versatile and synthesis approaches have been explored that solely start from input/output examples or that mine possible program fragments from public code repositories.
While the reported progress is impressive, we believe that the traditional deductive approach to synthesis has been unjustly neglected.

@@

The main goal of this project is to connect the classical approach to deductive synthesis --  proofs-as-programs~\cite{?}{ \color{red}s.o.}--, with modern automated theorem provers, in particular first-order solvers (and SMT solvers???), which have made impressive progress over the recent years (as evidenced by yearly competitions\footnote{\url{http://www.tptp.org/CASC/}}).

We believe this approach offers the following advantages:
1) First-order solvers are particularly suitable to solving \emph{quantified} formulas as needed for synthesis.
2) Proofs-as-programs represents a longstanding approach to program extraction.
That is to say, the groundwork for program extraction from proofs has been laid, we just need to adapt this theory to the specifics of modern automated theorem provers. {\color{red}in the classical case not everything has been said, perhaps "much groundwork..." and "...adapt and extend..."}
3) We are able to take advantage of the highly sophisticated and optimized proof search strategies of first-order solvers instead of needing to define and implement \emph{ad-hoc} search strategies.
%Moreover, we will profit from any future advancements in automated theorem proving.

@@

We now give the details on (WP1):
An optimal procedure for synthesis would consist of
\begin{itemize}
\item transforming the specification into an alternate form that would guarantee the use of as few non-intuitionistic rules as possible and applicability of Friedman translation in as many places as possible
\item applying the prover to this modified specification obtaining a classical proof
\item apply Friedman translation or similar translation procedures in as many places as possible
\item extracting a program via Curry-Howard correspondence, applying control operators to non-intuitionistic remnants as necessary.
\end{itemize}
All of these elements except the initial modification of the specification already exist separately and in model settings such as arithmetic. 
Extending these results and combining them will be the first major undertaking of this project. 
This is also a necessary precondition for determining what initial modifications should be undertaken. 
We expect that for most real-world synthesis problems our procedure will be sufficient to transform classical proofs into usable and verified programs.

Finally we want to implement a real system on top of Vampire realizing the synthesis problem (WP2) which will target a suitable functional programming language.

Finally we suggest a third approach which has not been covered in literature yet, that is to pass a modified formula to the classical prover which can give us additional information about what an intuitionistic proof can look like.

We will investigate how to restrict proof search in order to steer the search towards programs from certain syntactic fragments, we will experiment with different axioms that can be given as input to the automated theorem provers and we will examine input/output pairs to the tool.

@@

>>>>

\section{Introduction, background and state of the art (max. 2500 characters)}
%What is the topic and broad background of this research project? What is the state of the art in this field?
Programs are usually first written and then tested/verified.
In contrast, program synthesis allows a system developer to derive a program in one step.
At the same time, program synthesis holds the promise of freeing the programmer from low-level details and of enabling the programmer to focus on the high-level intent.
Program synthesis is a long-standing problem that crosses many areas of computer science and mathematics including programming languages, AI, formal methods and proof theory.
Traditionally, program synthesis has been formalized as a problem in deductive theorem proving~\cite{conf/ijcai/MannaW79}:
Given a logical formula that states that for all inputs there exists an output such that the required specification holds, one searches for a constructive proof of this formula and then extracts a program from this proof.
A classic route to program extraction is the \emph{Curry-Howard correspondence}~\cite{girard1989proofs} (also known as \emph{proofs-as-programs}) where one exploits the equivalence between proofs in intuitionist logic and functional programs in order to directly extract a functional program from an intuitionistic proof.
The Curry-Howard correspondence denotes a fundamental relationship between two seemingly unrelated formalisms, namely proofs systems on one hand and computational models on the other hand.
The Curry-Howard correspondence has been a major driving force for research spanning functional programming languages, type theory, proof systems and constructive logic.
However, progress on the deductive synthesis problem has been slow for many years because of a lack of automated and scalable provers.
Instead, a recent variant of the program synthesis problem has enabled dramatic progress (we refer to the special issue~\cite{fisman2022special} and the workshop\footnote{\url{https://simons.berkeley.edu/workshops/tfcs2021-1}} for an overview):
%\footnote{https://www.cs.technion.ac.il/~shaull/SYNT2022/index.html}
The logical specification is accompanied by a \emph{syntactic template} that constrains the space of possible implementations.
The synthesis procedure then searches for/enumerates candidate implementations and either rejects these candidates by finding violating input/output pairs or accept the generated program by verifying their correctness.
The syntactic approach is very versatile and synthesis approaches have been explored that solely start from input/output examples or that mine possible program fragments from public code repositories.
While the reported progress is impressive, we believe that the traditional deductive approach to synthesis has been unjustly neglected.

@@

The main goal of this project is to connect the classical approach to deductive synthesis --  proofs-as-programs~\cite{girard1989proofs} --, with modern automated theorem provers, in particular first-order solvers (and SMT solvers???), which have made impressive progress over the recent years (as evidenced by yearly competitions\footnote{\url{http://www.tptp.org/CASC/}}).

We believe this approach offers the following advantages:
1) First-order solvers are particularly suitable to solving \emph{quantified} formulas as needed for synthesis.
2) Proofs-as-programs represents a longstanding approach to program extraction.
That is to say, much groundwork for program extraction from proofs has been laid, we just need to adapt and extend this theory to the specifics of modern automated theorem provers.
3) We are able to take advantage of the highly sophisticated and optimized proof search strategies of first-order solvers instead of needing to define and implement \emph{ad-hoc} search strategies.
%Moreover, we will profit from any future advancements in automated theorem proving.

@@

We now give details on (WP1).
We will consider two approaches from the literature:

1) Parigot's $\lambda\mu$-calculus~\cite{Parigot1} gives a direct translation of classical proofs with datatypes and is essentially ready for implementation.
However, since non-intuitionistic tautologies like $\neg\forall xA(x)\to \exists x\neg A(x)$ or $A\vee \neg A$ are assigned a program, we lose the realizability of $\exists$ and $\vee$.
{\color{red} unclear: ``are assigned a program''}
As discussed in~\cite{practical}, in practice it is often possible to eliminate the non-intuitionistic fragments.
{\color{red} unclear: ``fragments''}
We aim to establish stronger theoretical guarantees on when this is possible but also more practical benchmarks on the feasibility of this approach, even when realizability is not ensured by a theorem.
{\color{red} unclear: ``more practical benchmarks''}

2) Sometimes classical validity guarantees intuitionistic validity, most famously for $\Pi_2$-formulas in Arithmetic~\cite{Friedman}. There have been various variations of this result, most notably by Schwichtenberg's group~\cite{schwichtenberg}.
We will adapt their work to our more general setting.
{\color{red} Can the previous sentence be made more concrete? Also why we this promising and feasiable?}
%	
%We expect that in practice these approaches will be sufficient in many situations.

Third, we also investigate a completely new direction:
We aim for a translation between provers rather than proofs. 
E.g., we study under what conditions classic proofs of a modified input formula allow to extract intuitionistic proofs of the actual input.
A promising example is actually the reverse direction:
given an intuitionistic prover one can establish classical provability via double negation translation.
%A first approximation has been outlined in~\cite{RDNT}.

Using the results of (WP1) we will implement a synthesis system on top of Vampire (WP2), targeting a suitable functional programming language.
We will experiment with various benchmark problems (e.g., median computation, sorting algorithms, arithmetic functions such as sqrt) and compare our results with the SyGuS-competition.
We will use ideas from syntax-guided synthesis (sketched in 1) in order to restrict and guide proof search:
We will experiment with different formulations of arithmetic axioms, induction axioms and redundant axioms.
We will further examine if proof search is possible when only input/output pairs are given.

@@

Programs that are correct by construction represent a longstanding goal of computer science.
Correct Their is holds particularly true for safet

Due to the focus on basic research and the mathematical and formal methodology, the proposed research does not entail ethical aspects beyond the usual mandate of good scientific practice.
Similarly, the proposed research does not entail sex-specific or gender-related aspects.

\section{Further Needed Stuff}
